---
title: "2023: Day 4"
date: 2023-12-4
categories:
  - R
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/4)

[My data](input){target="_blank"}

## Part 1

There are a bunch of cards, which get points based on how many winning numbers do they have.
I have to calculate how many points are the cards worth *in total*.

```
Card 1: 41 48 83 86 17 | 83 86  6 31 17  9 48 53
Card 2: 13 32 20 16 61 | 61 30 68 82 17 32 24 19
Card 3:  1 21 53 59 44 | 69 82 63 72 16 21 14  1
Card 4: 41 92 73 84 69 | 59 84 76 51 58  5 54 83
Card 5: 87 83 26 28 32 | 88 30 70 12 93 22 82 36
Card 6: 31 18 13 56 72 | 74 77 10 23 35 67 36 11
```

Numbers to the right of the vertical bar |  are the winning numbers.
The first winning number to the left of the bar makes the card worth 1 point, and each sucesive winning number doubles the points the card is worth.

```{r boilerplate}
#| message=FALSE
library(tidyverse)
library(here)
```

Reading the input as a vector
```{r}
input <- read_lines(here("2023/day/4/input"))
input[1:10]
```

Trying to parse the input as tidy data:
```{r}
card_data <- tibble(
  input = input
) %>%
  separate(
    input,
    into = c("id", "data"),
    sep = ": "
  ) %>%
  mutate(
    id = as.integer(str_extract(id, "\\d+"))
  )

card_data
```

```{r}
card_data_split <- card_data %>%
  separate(
    data,
    into = c("my_numbers", "winning_numbers"),
    sep = " \\| "
  )

card_data_split
```

Creating a function that parses the sequences of numbers as numeric vectors:
```{r}
to_numbers <- function(x) {
  str_extract_all(x, "\\d+") %>%
  map(as.numeric)
}
```


Function that takes two vectors and calcultes how many elements from the first one are in the second one:
```{r}
calculate_n_winning <- function(x, y) {
  keep(x, ~magrittr::is_in(., y)) %>%
  length()
}
```


```{r}
card_data_processed <-
  card_data_split %>%
  mutate(across(c(my_numbers, winning_numbers), to_numbers)) %>%
  mutate(n_winning = map2_dbl(my_numbers, winning_numbers, calculate_n_winning))

card_data_processed
```

Now I need a function that doubles its output for each integer greater than 1. Specifically, the function should behave as follows:

- f(0) = 0
- f(1) = 1
- f(2) = 2
- f(3) = 4
- f(4) = 8

```{r}
calculate_points <- function(x) {
  if (x == 0) {
    return(0)
  } else {
    return(2^(x-1))
  }
}
```

Applying the function to calculate how many points the cards are worth:
```{r}
card_data_processed %>%
  mutate(points = map_dbl(n_winning, calculate_points)) %>%
  count(wt = points)
```

The solution is correct 🥳

## Part 2

Here I think I need to use some form of recursion:

> "you win copies of the scratchcards below the winning card equal to the number of matches. So, if card 10 were to have 5 matching numbers, you would win one copy each of cards 11, 12, 13, 14, and 15."

> "Copies of scratchcards are scored like normal scratchcards and have the same card number as the card they copied. So, if you win a copy of card 10 and it has 5 matching numbers, it would then win a copy of the same cards that the original card 10 won: cards 11, 12, 13, 14, and 15. This process repeats until none of the copies cause you to win any more cards. (Cards will never make you copy a card past the end of the table.)"

It seems that the only column I need is `n_winning`, along with a new column to keep track of how many copies of the cards I have. I'll store the data in a matrix to access and modify it using indices.

```{r}
card_data_pt2 <-
  card_data_processed %>%
  transmute(
    n_matches = n_winning,
    count = 1
  ) %>%
  as.matrix()

card_data_pt2[1:10]
```

Now I need to code a procedure that captures the logic described in the prompt:
```{r}
#| cache: true
n_original_cards <- nrow(card_data_pt2)

for (i in 1:n_original_cards) {
  n_matches_i <- card_data_pt2[i, 1]
  count_i <- card_data_pt2[i, 2]

  if (n_matches_i == 0) next

  # If I have N copies of the current card, I'll win copies of the cards below N times
  for (n in 1:count_i) {
    for (j in 1:n_matches_i) {
      if (i+j <= n_original_cards) {
        # Adding one copy of each of the j cards below
        card_data_pt2[i+j, 2] = card_data_pt2[i+j, 2] + 1
      }
    }
  } 
}
```

Adding up the scratchcards using the `count` column:
```{r}
sum(card_data_pt2[,2])
```